---
title: "Group 1: Tennis "
author: "Olivia Beck, Emma Wilson, Ryan Volkert"
date: "December 12, 2019"
output: beamer_presentation
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("funs.R")
library(knitr)
library(kableExtra)
library(tidyverse)
set.seed(400)
```

## The Game of Tennis 

- Point: Smallest unit of measurement (Love-15-30-40-game)

- Game: A game is one when a player reaches 4 points with at least a 2 point advantage 

- Set: A set consists of 6 games and is won by the player who reaches 6 games first 

- Advantage Set: If a game score of 6-6 is reached and advantage set rules are used, a player can only win a set with a 2 game lead

- Matches: Best of 3 set (for women) or 5 sets (for men)

- Tie-break game: If a game score of 6-6 is reached and tie-break set rules are used. In a tie-break game, a player/team must reach 7 points with a two point advantage to win




## Values

Things we need to know:

- Empirical probability of winning a rally on serve 

- Empirical probability of winning a game on serve  \pause


Things we need to calculate: 

- Theoretical probability of winning a game on serve

- Theoretical probability of winning a set

- Theoretical probability of winning a match

- Theoretical probability of winning the tie-breaker 

## Slide with R Output

```{r, echo = FALSE}

TDWomen <- as.data.frame(matrix(0, nrow = 4, ncol = 3))

colnames(TDWomen) <- c("P(Win a Rally)", "Empirical P(Win a Game)", "Paper P(Win a Game)" )

TDWomen[1, ] <- c(0.69, 0.71, 0.89)
TDWomen[2, ] <- c(0.63, 0.80, 0.79)
TDWomen[3, ] <- c(0.65, 0.85, 0.83)
TDWomen[4, ] <- c(0.63, 0.77, 0.79)

OurDatWomen <- as.matrix(round(c(pg(TDWomen[1,1]), pg(TDWomen[2,1]), pg(TDWomen[3,1]), pg(TDWomen[4,1])),2))

TDWomen$`Our P(Win a Game)` <-(OurDatWomen)

kable(TDWomen, format = "latex", booktabs = TRUE)%>%
          kable_styling(latex_options = "scale_down")

```

## Slide with Plot

```{r pressure}
plot(pressure)
```

