---
title: "Group 1: Tennis "
author: "Olivia Beck, Emma Wilson, Ryan Volkert"
date: "December 12, 2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Game of Tennis 

- Point: Smallest unit of measurement (Love-15-30-40-game)

- Game: A game is one when a player reaches 4 points with at least a 2 point advantage 

- Set: A set consists of 6 games and is won by the player who reaches 6 games first 

- Advantage Set: If a game score of 6-6 is reached and advantage set rules are used, a player can only win a set with a 2 game lead

- Matches: Best of 3 set (for women) or 5 sets (for men)

- Tie-break game: If a game score of 6-6 is reached and tie-break set rules are used. In a tie-break game, a player/team must reach 7 points with a two point advantage to win




## Values

Things we need to know:

- Empirical probability of winning a rally on serve 

- Empirical probability of winning a game on serve  \pause


Things we need to calculate: 

- Theoretical probability of winning a game on serve

- Theoretical probability of winning a set

- Theoretical probability of winning a match

- Theoretical probability of winning the tie-breaker 

## Slide with R Output

```{r, echo = TRUE}





```

## Slide with Plot

```{r pressure}
plot(pressure)
```

